Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-14 doc/invt.tex
small edits to invt.tex

The Wang-Landau (WL) algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential.
%
%The bias potential is associated with discrete values of $z$.
%
The difference lies
in the updating window function,
which specifies
a neighborhood around the current value of $z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
The window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
and hence that of the PMF\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, the optimal updating schedule
is a function of the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of simulation steps.
\note{
  About the origin of the inverse-time formula.
  Besides the Belardinelli paper,
  we can also find it in Eq. (6) of reference \cite{liang2007},
  if I am not too mistaken.
  But I find the paper is a bit hard to read.
  There, however, the formula is $t_0/t$ instead of $1/t$.
}
%
The optimal schedule for metadynamics,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
As the mass is defined from the updating window function,
the resulting optimal schedule
is different in the WL and metadynamics cases.
%
While the optimal schedule in the WL case
is given by the known inverse-time formula,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
In either case, however,
the initial updating magnitude
can be optimally set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time in the WL case.
%
Further, we can generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-shifted inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

This is a functional of the curve $q(t)$.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process,
while the second term is the action of a free particle,
which must move at a constant velocity, $c$:
%
%We will further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
%By the Euler-Lagrange equation, we get
%$$
%\begin{aligned}
%0
%&=
%\frac{d}{dt} \int_0^T
%  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
%\\
%&= \int_0^T
%  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
%\end{aligned}
%$$
%where we have dropped the boundary terms
%by assuming Eq. \eqref{eq:kappa_limit}
%and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
%In order for this to hold for any $t$,
%we must have
%%
%$
%\ddot u\bigl( q(t') \bigr) = 0
%,
%$
%or
\begin{equation}
  \dot u\bigl( q(t') \bigr) = c
  .
\label{eq:dudt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:u_def}, we have
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$,
we get Eq. \eqref{eq:alpha_invt1}.

@@

The optimization procedure is similar to the single-bin case.
%
However, we will assume some restrictions and approximations
to simplify the discussion.

@@

The matrix, $\mathbf w$, is not arbitrary.
%
First, to sample
the desired distribution,
$\mathbf p = (p_1, \dots, p_n)$,
${\dot V}_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant of $i$
if $h_j(t)$ were the same as $p_j$.
%
This allows the bias potential to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}.

To simplify the ensuing discussion,
we  limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e. $\mathbf \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

@@

\subsubsection{\label{sec:eqlerr}
Equilibrium error under a constant updating magnitude
}

>>>>

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012},
more commonly used for MC and MD simulations, respectively\cite{junghans2014}.
%
%flat-distribution-sampling (FDS) techniques
%that build the bias potential adaptively or on the fly.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential,
%and they are closely-related\cite{junghans2014, micheletti2004}.
%
%The bias potential is associated with discrete values of $z$.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of $z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Usually, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
Although the assignment of the window functions to the two methods
is not absolute\cite{micheletti2004, kim2006, *kim2007, junghans2014},
below we will use the terms WL and metadynamics
to represent FDS simulations using the
$\delta$- and Gaussian window functions, respectively.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
and hence that of the PMF\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, the optimal updating schedule
is a function of the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of simulation steps.
\note{
  About the origin of the inverse-time formula.
  Besides the Belardinelli paper,
  we can also find it in Eq. (6) of reference \cite{liang2007},
  if I am not too mistaken.
  But I find the paper is a bit hard to read.
  There, however, the formula is $t_0/t$ instead of $1/t$.
}
%
%The optimal schedule for metadynamics,
%however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
with a general window function,
including the $\delta$- and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
As the mass is defined from the updating window function,
the resulting optimal schedule
is different in the WL and metadynamics cases.
%
While the optimal schedule in the WL case
is given by the known inverse-time formula,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
In either case, however,
the initial updating magnitude
can be optimally set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time in the WL case.
%
Further, we can generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-shifted inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

This is a functional of the curve $q(t)$.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process,
while the second term is the action of a free particle,
which must move at a constant velocity, $c$:
%
%We will further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
%By the Euler-Lagrange equation, we get
%$$
%\begin{aligned}
%0
%&=
%\frac{d}{dt} \int_0^T
%  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
%\\
%&= \int_0^T
%  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
%\end{aligned}
%$$
%where we have dropped the boundary terms
%by assuming Eq. \eqref{eq:kappa_limit}
%and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
%In order for this to hold for any $t$,
%we must have
%%
%$
%\ddot u\bigl( q(t') \bigr) = 0
%,
%$
%or
\begin{equation}
  \dot u\bigl( q(t') \bigr) = c
  .
\label{eq:dudt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:u_def}, we have
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$,
we get Eq. \eqref{eq:alpha_invt1}.
%
From the boundary condition at $t' = T$,
we get $c = 1/(T + t_0)$,
and from the case of $t' = 0$,
we get $q(T) = \ln(1 + T/t_0)$.
%
Since the above optimization is performed
under a fixed $q(T)$, below
one will choose a proper value of $q(T)$,
or, equivalently, of $t_0$, to complete the optimization.

@@

The optimization procedure is similar to the single-bin case.
%
The key simplification comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$.
%However, we will assume some restrictions and approximations
%to simplify the discussion.

@@

The matrix, $\mathbf w$, is subject to three conditions.
%
First, to sample
the desired distribution,
$\mathbf p = (p_1, \dots, p_n)$,
${\dot V}_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant of $i$
if $h_j(t)$ were the same as $p_j$.
%
This allows the bias potential to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}.
%
The quoted results were unaffected by the allowance
of negative elements into the matrix, $\mathbf w$.

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e. $\pmb \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.
%
In the way,
the error weighting distribution, $\pmb \rho$,
restrains the form of the updating matrix, $\mathbf w$.
%
The condition, albeit somewhat artificial,
is readily met in many updating schemes used in practice:
most commonly, $\pmb \rho$ is a uniform distribution,
and this condition requires the updating matrix, $\pmb w$, to be symmetric.

@@

\subsubsection{\label{sec:eqlerr}
Modeling initial error from a simulation
under a constant updating magnitude
}
